Bayesian exponential random graph modeling of KWL Burial network

### 2020-12-04 analysis/paper/paper.Rmd
some edits to the discussion section

<!-- comparison between two networks using patchwork bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand if this network size difference affects the posterior distributions, we conducted a sensitivity test by randomly removing `r diff_pre_post ` nodes from the post-European network to resemble the size of pre-European network with vertex bootstrap. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence interval for the network population, especially for smaller networks. We set the resampling size to 1,000 networks to compute network statistics, including density, clustering, and degree, for two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network density and degree, but higher extreme network clustering. This pattern is consistent before and after node removal for the post-European network, and the difference can be still identified between networks from different phases. This indicates our networks are relatively robust to differences in sample size and suggests a very small sample size effect.

Similarly, we can use vertex bootstrap to compare pre- and post-European networks since it allows statistical testing for multiple networks of different sizes [@Chen2019; @Snijders1999]. We used t-test for simulated networks and the results show that there are significant differences in all network statistics from both periods. In addition, since the prior setting will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

@@

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path [@Hunter2008]) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falls outside the 95% interval, the fitting is generally good with most observations within it. This also suggests that our observed network statistics do not follow a normal distribution, especially for the network after the arrival of the Europeans that presents a bimodal pattern. We also compared the first three distribution moments of each observed distribution and their corresponding simulated distribution represented by mean. The figure shows that modeled values from the pre-European network are closer to the observed values overall, compared to the values from the post-European network. For the variance of degree, the values from modeled post-European distribution are closer to the observed distribution. In general, this demonstrates a better fit for the pre-European network. However, it should be noted that the differences between the statistics of two models are relatively small, which still allows direct comparison of two networks.

<!-- theoretical framework for the results-->
The notable finding in our results is the change in network properties of Kiwulan burials from a more cohesive network (represented by higher transitivity and negative popularity) toward a more centralized network (lower transitivity and positive popularity) after the European presence during the 17th century. This confirms our hypothesized models of chronological differences in the structure of burials in terms of grave goods and burial attributes associated with demography and mortuary practices. The differences are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum, that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000]. Compared to vertical hierarchy that stresses stratified social characteristics, the corporate-network model is more effective to investigate the way how a small scale society organized and explore changes in social relations in a pericolonial context. A corporate-based society stresses shared power between individuals, integrative rituals, and social inequality, if any, would be associated with a social group or sector [@Siegel1999]. In contrast, a network-based society presents wealth accumulation through individual networks, resource control, and trade monopolization.

<!--archaeological implication based on the results-->
The covariants for both networks before and after the European presence provide further information of which variables relative to the formation of networks with the same prestige goods in common. We included age, sex, ritual pottery, and wealth index based on the value of grave goods. The pre-European network indicates that burials with pottery as ritual vessels tend to be tied, while the burials within the same level of wealth tend to be connected in the post-European network. Moreover, the burials with ritual vessels tend not to be tied in the post-European network. This illustrates there are different mechanisms to form relations between burials in terms of access to trade goods. Before the European arrival, ritual behavior seem to be an emphasized cultural practice relative to the use of prestige goods. After the European arrival, the cultural emphasis shifted from ritual to wealth difference. The demographic information, such as age and sex, do not show any correlations with the network formation for both periods. This may be influenced by some unidentified burials that limits a more insightful result. The distance between burials is not a variable that correlates to the network formation either.

>>>>

<!-- comparison between two networks using patchwork bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand if this network size difference affects the posterior distributions, we conducted a sensitivity test by randomly removing `r diff_pre_post` nodes from the post-European network to resemble the size of pre-European network using a vertex bootstrap procedure. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) of a network to quantify standard errors in the network statistics of interest. This enables the evaluation of uncertainty for networks and the comparison between networks by examining their confidence intervals for the network population, especially for smaller networks. We set the resampling size to 1,000 networks to compute network statistics, including density, clustering, and degree, for our two networks. The result shows that the 95% confidence interval of each network statistic contains the observed network density and degree, but the observed network clustering values are higher than the 95% CI of the bootstrap distributions. This pattern is consistent before and after node removal for the post-European network. The bootstrap distributions for the post-European network, when adjusted to be the same size pre-European network, show distinct differences from the pre-European network statistics. This indicates our networks are relatively robust to differences in sample size, and suggests a negligible sample size effect.

Similarly, we can use vertex bootstrap to compare pre- and post-European networks since it allows statistical testing for multiple networks of different sizes [@Chen2019; @Snijders1999]. We used a t-test on the distributions of network statistics from the bootstrapped network samples. The results show that there are significant differences in all network statistics from both periods. This supports the presence of anthropologically meaningful differences in the pre- and post-European networks.

<!-- move this text to another section that is a more logical fit --> 
In addition, since the prior settings will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path [@Hunter2008]) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falling outside the 95% interval, the fit is generally good with most observations within it. This also suggests that our observed network statistics do not follow a normal distribution, especially for the network after the arrival of the Europeans that presents a bimodal pattern. We compared the first three distribution moments of each observed distribution and their corresponding simulated distributions, represented by means. Figure \@ref(fig:xxxx) shows that modeled values from the pre-European network are closer to the observed values overall, compared to the values from the post-European network. For the variance of degree, the values from the modeled post-European distribution are closer to the observed distribution. In general, this demonstrates a better fit for the pre-European network. However, it should be noted that the differences between the statistics of two models are relatively small, which still allows direct comparison of two networks.

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network (represented by higher transitivity and negative popularity) toward a more centralized network (lower transitivity and positive popularity) after the European presence during the 17th century. This supports our hypothesized models of chronological change from a more corporate society to a more networked society, as indicated in the grave goods and burial attributes. These observations are in line with theoretical assumptions of horizontal hierarchy, the corporate-network continuum that focuses on distinct strategies for achieving power as an alternative explanation to understand the emergence and level of social inequality [@Drennan2010; @Feinman2000]. Compared to traditional models of vertical hierarchies that stress stratified social characteristics, the corporate-network model may be more relevant for understanding how a small scale society was organized, such as we see here in the pericolonial context of Kiwulan. A corporate-based society, such the pre-European situation at Kiwulan, stresses shared power between individuals, integrative rituals, and social inequality, if any, would be associated with a social group or sector [@Siegel1999]. In contrast, a more network-based society, indicated by the post-European data from Kiwulan, presents wealth accumulation through individual networks, resource control, and trade monopolization.

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). We included age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences.

Distance between burials did not correlate to the network formation... <!-- anthropological concept here -->

Demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of ge and sex values for many burials in our sample.

@@
